---
title: "EM algorithm comparison"
output: html_notebook
---

# EM algorithm comparison

We will now see whether the EM algorithm can resolve at least some
of the spurious mapping. For that we will compare it to the standard method
of calling mapping by the best alignment score and/or mapping score.

We start by reading our count data for both methods.

```{r}
library(mbtools)
library(ggplot2)
theme_set(theme_minimal())

counts <- fread("nanopore_counts.csv")
counts <- counts[sample %in% c("barcode01", "barcode05")]
head(counts)
```

We could just compare how many unique references were hit but in general we will
also apply abundance cutoffs to our data, so let us count the number of unique references by cutoff.

```{r}
cutoffs <- 10 ^ seq(-1, 3, length.out = 16)
nref <- lapply(cutoffs, function(co) {
    nr <- counts[counts > co, .N, by = c("method", "sample")]
    nr[, "cutoff" := co]
    nr
}) %>% rbindlist()
head(nref)
```

Now let's visualize that.

```{r}
ggplot(nref, aes(x = cutoff, y = N, color = method)) + 
    geom_point() + geom_line() + scale_x_log10() +
    facet_wrap(~ sample) + 
    labs(x = "abundance cutoff", y = "no. unique references")

ggsave("figures/algorithms.png", width = 8, height = 3)
```

As we can see the EM algorithm can resolve spurious mappings well for low
abundance where it reduces the spurious mappings by about half. For higher abundances 
both methods work equally well since there is sufficient coverage to resolve the
correct 16S reference.